Experience Report: Haskell in Computational Biology

### 2012-06-30 mrfy_experience_report.tex
Many changes. Ugly plot included, need help to fix.

Our Haskell code for hidden Markov models and Viterbi's algorithm
solves the same problems as existing C++~code.
\remark{Drop this \P\ or move it}
Other researchers using Haskell may also have to reimplement code,
but researchers may also reimplement code out of choice, not
necessity.
IN COMPUTATIONAL BIOLOGY, REIMPLEMENTING EXISTING ALGORITHMS IS UNREMARKABLE.
For~example, both SMURF and HMMER contain new implementations of these
same algorithms.
In~MRFy, we are using the Haskell implementations to help solve new
problems.

@@

Despite these non-modular aspects, the search
interface provides ample scope for
experiments.
\remark{This \P\ needs revising in light of the ``monad madness'' refactoring.}
Random hill climbing took fifty lines of code and one~day to
implement.
Simulated annealing reuses hill climbing's
 \texttt{nextPt}, \texttt{pt0}, and \texttt{quit} functions.
Simulated annealing's  new \texttt{accept} function took ten
lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm reuses the same
\texttt{pt0} and \texttt{quit} functions, and its \texttt{accept}
function took only a few lines of code.
But~the recombination of parent
placements in its \texttt{nextPt} function took forty lines of code
and a full day to implement.

\remark{Stop functions not as composable as we would like.
Wanted: combinator library.}

@@

We~also found an opportunity to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.
Even when there is just one placement per point,
MRFy can keep 6~cores busy.
\remark{This paragraph needs to talk about \textbf{speedups}.}
Given multiple placements per point,
it~can use 48~cores.
However, we have found that MRFy performs best, at least on our
hardware, with between 12 and 16 cores.
We have access to only
one 48-core system, and while MRFy will keep all 48 cores busy, it
performs only slightly better, at an average of 35 seconds to find an alignment
on a large\remark{QUANTIFY!!} protein structure, than when given 16
cores, which takes an average 
of 37 seconds.

MRFy's implementation of Viterbi's algorithm appears dramatically
slower than the C++ version implemented in \mrfy's predecessor,
SMURF.
On a 2.6GHz Intel Core i7,
given a contrived query sequence of 56,000 amino acids,
which is larger than any known protein,
with a hidden Markov model of 343 nodes and no beta strands,
\mrfy\ takes almost two minutes to compute an optimal parse.
SMURF parses the same sequence in half a second.
\remark{This result looks really bad. What is this contrived benchmark supposed to
mean, anyway??  Can we get a more realistic benchmark?
What fraction of time is spent in Viterbi?}
But unlike SMURF, \mrfy\ doesn't have to run on large models.
{\large{\textbf{QUANTIFY!}}}
\mrfy\ repeatedly runs Viterbi's algorithm on 
a~great many smaller models.
{\large{\textbf{QUANTIFY!}}}
Here, performance is good enough that \mrfy\ can do its job.
Given a Markov random field with a 12-stranded ``beta sandwich''
structure,
\mrfy\ typically computes an alignment within a minute;\remark{with or
without parallelism?}
SMURF allocates over 16GB of memory
and does not terminate even after eight hours.
%%% Had we implemented the SMURF algorithm in Haskell, we believe that performance
%%% would be unacceptable.
%%% However, for the algorithm underlying MRFy, we are satisfied with the runtime
%%% performance of Haskell and GHC.
%
% Nonetheless, we expect that we can still improve MRFy's run-time performance
% with further profiling and tuning.  
% And we expect that if we cross the street, we will get to the other side.

\subsection{Awkward profiling and debugging}

@@

We've shown you some of the code we like, but we've also written 
code we dislike.
The~code we like least is what we've written to work around
obstacles we 
encountered in profiling and debugging.

GHC's profiler assigns costs not to functions but to ``cost
centers'' \cite{sansom-pj}.
In GHC~7.0, which we used 
for most of MRFy's development, 
cost-center annotations can be added automatically only
to top-level 
functions.
But most of our code is in nested functions.
We~annotated such functions manually,
but the annotations
made our code so ugly that we felt compelled to
remove them as soon as possible.
Without the annotations, however, GHC's call-site and allocation
profiles did not give us enough information to be useful.
We~miss valgrind, whose profiler gives useful information
without \emph{any} annotations.
(GHC~7.4 provides much
better profiling tools.
We~look forward to using them.)

@@

SUMMARY OF WHAT LITTLE WE HAVE LEARNED FROM QUICKCHECK
Have we tested that \verb+vee'+ returns a most likely sequence?
And didn't we find an inconsistency?

@@

@@

In~\secref{awkward-profiling} above, we~mention the awkwardness of the
cost-center annotations required for profiling.
But~we also found it difficult just to get the profiler working.
In~particular, profiling requires that all installed libraries be recompiled
with profiling support.
Using Haskell's Cabal tool, which manages installed libraries,
we~were not able to achieve such recompilations consistently.

>>>>

@@

Despite these non-modular aspects, the search
interface provides ample scope for
experiments.
Random hill climbing took fifty lines of code and one~day to
implement.
Simulated annealing reuses hill climbing's
 \texttt{nextPt} function.
Simulated annealing's  new \texttt{utility} function took ten
lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm reuses the same
\texttt{pt0} and \texttt{SearchStop} functions, and its \texttt{utility}
function took only a few lines of code.
But~the recombination of parent
placements in its \texttt{nextPt} function took forty lines of code
and a full day to implement.

We note that the \texttt{SearchStop} functions are not as composable as we
would like, and we would like to have a combinator library for searches.

@@

We~also found an opportunity to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.
We have access to a
one 48-core, 2.3GHz AMD Opteron 6176 system, on which we benchmarked \mrfy\ performing
a search on a large protein structure (an ``8-bladed beta-propeller'')
comprising 343 hidden Markov model nodes and 40 beta strands, and a query sequence
592 amino acids long.
Figure~\ref{multicore} illustrates the mean speedup \mrfy\ achieves as the number of
processing cores increases.
Optimal speedup appears to occur between 12 and 16 processors.

\begin{figure}[h!]
\centerline{\includegraphics[width=6cm]{mrfy_multicore_log}}
\caption{MRFy run-time performance with varying
numbers of processing cores on a large protein}\label{multicore}
\end{figure}

MRFy's implementation of Viterbi's algorithm appears dramatically
slower than the C++ version implemented in \mrfy\'s predecessor,
SMURF.
Both \mrfy\ and SMURF, when provided with a Markov random field containing no beta strands,
simply compute the optimal solution using Viterbi's algorithm.
For both \mrfy\ and SMURF, this task admits no parallelism.
On a 2.6GHz Intel Core i7,
given a micro-benchmark consisting of a contrived query sequence
of 2000 amino acids (which represents a large protein),
and a hidden Markov model of 343 nodes and no beta strands,
\mrfy\ takes 2.32 seconds to compute a Viterbi solution.
SMURF performs the same computation in 0.29 seconds.

But unlike SMURF, \mrfy\ doesn't have to run on large models.
For example, on the ``8-bladed propeller'' benchmark, the beta strands
break the Markov random field into 41 pieces, each of which is much smaller
than the original (typically 10 or fewer nodes rather than 343).
And, each \texttt{placement} breaks the query sequence into 41 pieces, each of
which is much smaller than the original (typically 20 or fewer amino acids
rather than 592).
Thus, \mrfy\ repeatedly runs Viterbi's algorithm on 
a~great many smaller models, and the independence of these models
means that \mrfy\ can execute them in parallel.
Here, performance is good enough that \mrfy\ can do its job.
Given a Markov random field with a 12-stranded ``beta sandwich''
structure,
\mrfy\ typically computes an alignment within a minute;
SMURF allocates over 16GB of memory
and does not terminate even after eight hours.

Clearly, micro-benchmarks do not tell the whole story.
Our goal is to solve a biological problem, and though our implementation of
Viterbi's algorithm is comparatively slow, we can still solve a biological problem
quickly.
%%% Had we implemented the SMURF algorithm in Haskell, we believe that performance
%%% would be unacceptable.
%%% However, for the algorithm underlying MRFy, we are satisfied with the runtime
%%% performance of Haskell and GHC.
%
% Nonetheless, we expect that we can still improve MRFy's run-time performance
% with further profiling and tuning.  
% And we expect that if we cross the street, we will get to the other side.

\subsection{Awkward debugging and testing}

@@

We've shown you some of the code we like, but we've also written 
code we dislike.
The~code we like least is what we've written to work around
obstacles we 
encountered in debugging.

@@

Based on the QuickCheck properties we did write, we have identified and
easily fixed a bug in our implementation of Viterbi's algorithm; we had not
been including the score for the transition from the final node of the hidden
Markov model to the special ``end'' state.
Without QuickCheck, it is unlikely that we would have found this bug,
and if we had, determining the cause would have required much time.

@@

Our Haskell code for hidden Markov models and Viterbi's algorithm
solves the same problems as existing C++~code.
Other researchers using Haskell may also have to reimplement code,
but researchers may also reimplement code out of choice, not
necessity.
In computational biology, reimplementing existing algorithms is unremarkable.
For~example, both SMURF and HMMER contain new implementations of these
same algorithms.
In~MRFy, we are using the Haskell implementations to help solve new
problems.

@@

GHC's profiler assigns costs not to functions but to ``cost
centers'' \cite{sansom-pj}.
In GHC~7.0, which we used 
for most of MRFy's development, 
cost-center annotations can be added automatically only
to top-level 
functions.
But most of our code is in nested functions.
We~annotated such functions manually,
but the annotations
made our code so ugly that we felt compelled to
remove them as soon as possible.
Without the annotations, however, GHC's call-site and allocation
profiles did not give us enough information to be useful.
We~miss valgrind, whose profiler gives useful information
without \emph{any} annotations.
GHC~7.4 provides much
better profiling tools, which eliminate most of these obstacles,
rendering this experience no longer as relevant.
